---
title: "README"
format: gfm
editor: visual
---

# Inferring processes within dynamic forest models using hybrid modeling

This repository contains the code to reproduce the results in "**Inferring processes within dynamic forest models using hybrid modeling**" by Maximilian Pichler & Yannek Käber

## Preprint

Preprint can be found on the [arXiv server](https://arxiv.org/abs/2508.01228)

## Data

Results are based on simulated data and on empirical data. The original BCI plot data is available from <https://doi.org/10.15146/5xcp-0d46> , meteorological data from <https://doi.org/10.15486/ngt/1771850>, water potential map from <https://doi.org/10.6084/m9.figshare.c.4372898.v1>, and species list from <https://doi.org/10.1126/science.aaz4797>. The FINN model documentation is provided in the SI Appendix. FINN is available as R package and can be installed from <https://github.com/FINNverse/FINN>.

## Scripts

Scripts to reproduce the analysis can be found in the `code` folder. To rerun everything, run in the following order:

Prepare data (from raw BCI data):

```{r, eval=FALSE}
source("code/data_preparation/01-bci-data-preparation.R")
source("code/data_preparation/01-uholka-data-preparation.R")
source("code/data_preparation/01b-bci-secundary-forest-preparation.R")
source("code/data_preparation/02-spatial-holdout-folds-bci.R")

```

Run simulations and models:

-   Run scripts from 01-07 in the code folder (GPU support is recommended!)

## Results

{{< include Results.qmd >}}

## Session info

```{r}
sessionInfo()
```
