---
title: ""
format: 
  docx:
    keep-md: false
    fig-format: pdf
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
  title-delim: ":"
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(dev = "cairo_pdf")
knitr::opts_chunk$set(fig.path="figures/", echo = FALSE)
```


```{r}
#| echo: false
#| message: false
library(ggplot2)
library(data.table)
source("code/plot-functions.R")
library(gridExtra)
library(grid)
library(FINN)
library(ggbeeswarm)
library(gghalves)
library(data.table)
library(tidyverse)
library(hexbin)

FINN::FINN.seed(1)

pft_cols <- c(
  Slow                         = "#BF7DA5",  # pink-mauve
  Fast                         = "#DBA242",  # ochre-orange
  `Long-lived pioneer (LLP)`   = "#479C77",  # medium green
  `Short-lived breeder (SLB)`  = "#3471AE",  # mid blue
  Intermediate                 = "#C36837"   # terracotta
)
pft_cols_int <- pft_cols
names(pft_cols_int) <- 1:5

# Function to extract the legend as a grob
get_legend <- function(ggplot_obj) {
  # Build the gtable
  plot_gtable <- ggplot_gtable(ggplot_build(ggplot_obj+guides(colour = guide_legend(nrow = 1))))
  
  # Find the index of the guide box (the legend)
  guide_index <- which(sapply(plot_gtable$grobs, function(x) x$name) == "guide-box")
  
  # Return the legend grob
  if (length(guide_index) > 0) {
    return(plot_gtable$grobs[[guide_index]])
  } else {
    warning("No legend found.")
    return(NULL)
  }
}

label_grob_f = function(x){
  textGrob(x, x = unit(0.05, "npc"), y = unit(0.95, "npc"), just = c("left", "top"), gp = gpar(fontsize = 16, fontface = "bold"))
}


#' Draw a GAM smooth (±SE) without plotting points
#'
#' @param x,y        numeric vectors (explanatory, response) of equal length.
#' @param n          grid size for predictions (controls curve resolution).
#' @param se.mult    SE multiplier (≈2 for 95 % band).
#' @param add        logical; overlay on current plot (`TRUE`) or start a new
#'                   blank frame (`FALSE`, default).
#' @param formula    GAM formula (default = y ~ s(x, bs = "cs")).
#' @param family     glm family; if NULL and all(y > 0) defaults to
#'                   `Gamma(link = "log")`, otherwise `gaussian()`.
#' @param gam_args   named list of extra arguments for `mgcv::gam()` (optional).
#' @param col,lwd    colour / line-width for the central smooth.
#' @param se.col,se.lty  colour / lty for the ±SE curves.
#' @param ...        further arguments passed **only** to `plot()` when
#'                   `add = FALSE` (axis labels, limits, …).
#'
#' @return invisibly a data.frame of x, fit, upr, lwr.
#'
#' @examples
#' set.seed(1)
#' x <- runif(120, 0, 10)
#' y <- exp(0.5 * sin(x)) + rlnorm(120, 0, 0.1)
#'
#' ## (1) default smooth chosen for positive data
#' lines_gam(x, y)
#'
#' ## (2) a *smoother* curve with REML + higher gamma
#' plot(x, y, pch = 19, col = "#00000044")
#' lines_gam(
#'   x, y,
#'   add       = TRUE,
#'   formula   = y ~ s(x, k = 30),     # more basis functions
#'   gam_args  = list(method = "REML", gamma = 1.6),
#'   col       = "tomato"
#' )
lines_gam <- function(x, y,
                      n        = 100,
                      se.mult  = 2,
                      add      = FALSE,
                      formula  = NULL,
                      family   = NULL,
                      gam_args = list(),
                      col      = "#6c9fca",
                      lwd      = 3,
                      se.col   = col,
                      se.lty   = 2,
                      se = F,
                      ...) {
  
  ## --------- sanity checks & setup -----------------------------------------
  stopifnot(length(x) == length(y))
  if (!requireNamespace("mgcv", quietly = TRUE))
    stop("Package 'mgcv' is required; install it with install.packages('mgcv').")
  
  dat <- data.frame(x = x, y = y)
  
  ## choose family if none supplied
  if (is.null(family))
    family <- if (all(y > 0)) Gamma(link = "log") else gaussian()
  
  ## default formula if none supplied
  if (is.null(formula))
    formula <- y ~ s(x, bs = "cs")
  
  ## --------- fit model (extra args allowed) ---------------------------------
  gam_call <- c(
    list(formula = formula, data = dat, family = family),
    gam_args
  )
  mod <- do.call(mgcv::gam, gam_call)
  
  ## --------- predictions on link scale -------------------------------------
  xs  <- seq(min(x, na.rm = TRUE), max(x, na.rm = TRUE), length.out = n)
  pr  <- predict(mod, newdata = data.frame(x = xs), se.fit = TRUE)
  
  fit.lin <- pr$fit
  se.lin  <- pr$se.fit
  
  ## back-transform to response scale
  linkinv <- family$linkinv
  fit <- linkinv(fit.lin)
  upr <- linkinv(fit.lin + se.mult * se.lin)
  lwr <- linkinv(fit.lin - se.mult * se.lin)
  
  ## --------- draw -----------------------------------------------------------
  if (!add) {
    plot(x, y, type = "n", ...)       # blank plotting region
  } else if (is.null(dev.list())) {
    stop("No active graphics device – call plot() first or set add = FALSE.")
  }
  
  lines(xs, fit, col = col,      lwd = lwd)
  if(se == T){
    lines(xs, upr, col = se.col,   lty = se.lty)
    lines(xs, lwr, col = se.col,   lty = se.lty)
  }
  
  invisible(data.frame(x = xs, fit = fit, upr = upr, lwr = lwr))
}
```



```{r}
#| echo: false
#| message: false

# Read fitted models

chosen_scale = "pft-period7-25patches"
fitted_models <- c(
    list.files("results/02_realdata/", full.names = T, recursive = T, pattern = chosen_scale),
    list.files("results/02_realdata_hybridTF0dropout0.1/", full.names = T, recursive = T, pattern = chosen_scale)
    # list.files("results/02_realdata_hybridSmall/", full.names = T, recursive = T, pattern = chosen_scale)
)
fitted_models <- fitted_models[grepl("S0_T0",fitted_models)]

```

## Main Text

### Figure 2 - Recovery of functional form

```{r}
#| label: fig-figure_2
#| fig-format: pdf
#| fig-width: 12
#| fig-height: 6
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Hybrid-FINN recovers the functional form of a process and improves predictive performance. To test whether Hybrid-FINN can recover the functional form of a process, we simulated data from a Process-FINN model (M1), in which the functional form of the growth process includes an environmental unimodal niche curve (quadratic effect of the environment). Then, we tested two models: M3, a Hybrid-FINN in which growth was replaced by a deep neural network; and M2, a misspecified Process-FINN model with an assumed linear relationship between environment and growth. (a) compares the calibrated growth-environment relationships of models M2 and M3 to the true model M1. (b) compares the predictive performance of models M2 and M3 with the observed data, measured by Spearman rank correlation. Correlations are based on equilibrium simulations and represent the relation for each variable and site over 500 simulated timesteps."

load("results/functional_form.RData")
formula = y ~ s(x,bs="cr",  k= 5)

par(mfrow = c(1, 2))

plot(pred_m1env_dt$env1, (pred_m1env_dt$growth*pred_m1env_dt$dbh), col = NA, ylab = "dbh growth", xlab = "environment", ylim = c(0, 5), las = 1)
mtext("a", side = 3, line = 0.1, adj = -0.3, font = 2, xpd = NA)  # 'a' at top-left
lines_gam(pred_m1env_dt$env1, (pred_m1env_dt$growth*pred_m1env_dt$dbh), col = "black", add = T, formula = formula, ylim = c(0,3))
lines_gam(pred_m2env_dt$env1, (pred_m2env_dt$growth*pred_m2env_dt$dbh), col = "#fe944e", add = T, formula = formula, ylim = c(0,3))
lines_gam(pred_m3env_dt$env1, (pred_m3env_dt$growth*pred_m3env_dt$dbh), col = "#6c9fca", add = T, formula = formula, ylim = c(0,3))
legend("topleft", legend = c("M1 (true form)", "M2 (wrong form)", "M3 (hybrid model)"), col = c("black", "#fe944e", "#6c9fca"), lty = 1, bty = "n", lwd = 3)
text(-2.5, 5.40, xpd = NA, labels = "a", font = 2, cex = 1.4 )


boxplot(r~model+variable2, data=cor_dt_form, xaxt = "n", las = 1, at = c(0, 1, 3, 4, 6, 7, 9, 10, 12, 13, 15, 16), col = c("#fe944e", "#6c9fca"), xlab = "")
axis(1, at= c(0.5, 3.5, 6.5, 9.5, 12.5, 15.5), labels = rep("", 6))
text(x = c(0.5, 3.5, 6.5, 9.5, 12.5, 15.5), levels(cor_dt_form$variable2), y = -0.42, xpd = NA)
legend("bottomleft", bty = "n", legend = c("M2 (wrong from)", "M3 (hybrid model)"), col = c("#fe944e", "#6c9fca"), pch = 15)
text(-2.5, 1.11, xpd = NA, labels = "b", font = 2, cex = 1.4 )

```


### Figure 3 - Spearman correlation between predicted and observed

```{r}
#| label: fig-figure_3
#| fig-format: pdf
#| fig-width: 7
#| fig-height: 5
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Spearman correlation of observed and simulated variables for five PFTs from the five-fold blocked spatial cross-validation of all models calibrated/trained on the BCI data. 'Process' refers to Process-FINN as a full process model with fixed functional forms, 'Hybrid' refers to Hybrid-FINN in which the growth process is replaced by a DNN, 'Naïve NN' refers to a single DNN that was trained on the response variables for each time step. The naïve NN received the derived response variables of each previous time step as well as the current environment as input. PFTs are according to (Rüger et al., 2020). For a similar figure with the RMSE see Figure A1."

chosen_scale = "pft-period7-25patches"

fitted_models_fig4 <- fitted_models

## create fit simulations
fit_process = build_model_dt(pt_file = fitted_models_fig4[1])[[1]]
i_hybrid_model <- fitted_models_fig4[grepl("hybrid",fitted_models_fig4)][1]
for(i_hybrid_model in fitted_models_fig4[grepl("hybrid",fitted_models_fig4)]){
  
  fit_hybrid = build_model_dt(pt_file = i_hybrid_model)[[1]]
  all_dt = rbind(
    data.table(fit_process[test_train == "train"], type = "process"),
    data.table(fit_hybrid[test_train == "train"], type = "hybrid")
  )
  period_length = unique(all_dt[!is.na(period_length)]$period_length)
  all_dt <- all_dt[,-c("period_length")]
  
  all_dt2 <- melt(all_dt, id.vars = c("type","siteID","year","species","test_train"))
  all_dt2[grepl("pred",variable), pred_obs := "pred",]
  all_dt2[grepl("obs",variable), pred_obs := "obs",]
  all_dt2[, variable := gsub(".obs|.pred","",variable),]
  
  p_dat <- all_dt2[test_train == "train",.(value = mean(value,na.rm = T)), by = .(type,year,species,test_train, pred_obs, variable)]
  p_dat2 <- p_dat[!is.na(value)]
  
  p_dat2_all <- 
    rbind(
      p_dat2[variable %in% c("ba", "trees"),.(
        value = sum(value, na.rm = T),
        species = "all"
      ), by = .(type, year, test_train, pred_obs, variable)],
      p_dat2[variable %in% c("dbh", "reg","growth","mort"),.(
        value = mean(value, na.rm = T),
        species = "all"
      ), by = .(type, year, test_train, pred_obs, variable)]
    )
  
  p_dat2 <- rbind(
    p_dat2,
    p_dat2_all
  )
  
  p_dat2[variable %in% c("ba", "trees"), value := value/0.1,]
  p_dat2[variable %in% c("mortality", "growth"), value := value/5*100,]
  
  cors_fullPlot = fread("results/cors_fullPlot.csv")
  cors_fullPlotSpecies <- fread("results/cors_fullPlotSpecies.csv")
  
  all_cors <- rbind(
    data.table(cors_fullPlot, species = "all"),
    cors_fullPlotSpecies
  )
  p_dat2[,species2 := factor(species, levels = 1:5, labels = names(pft_cols)),]
  
  all_cors <- all_cors[scale == chosen_scale & grepl("T0", cv),]
  all_cors[data == strsplit(i_hybrid_model,"/")[[1]][2],type := "hybrid",]
  all_cors[data == "02_realdata",type := "process",]
  all_cors[, type := factor(type, levels = c("process", "hybrid"), ordered = T),]
  all_cors <- all_cors[!is.na(type)]
  all_cors[,species := factor(species, levels = c(sort(unique(all_cors$species)))),]
  all_cors[,species2 := factor(species, levels = 1:5, labels = names(pft_cols)),]
  
  vars = c("ba", "trees", "dbh", "reg","growth","mort")
  var_labels <- c(
    expression(basal~area~(m^2~ha^-1)),
    expression(trees~(ha^-1)),
    expression(dbh~(cm)),
    expression(regeneration~(ha^-1)),
    expression(growth~(m^3~ha^-1~yr^-1)),
    expression(mortality~(m^3~ha^-1~yr^-1))
  )
  var_labels2 <- c(
    "basal~area~(m^2~ha^-1)",
    "trees~(ha^-1)",
    "dbh~(cm)",
    "regeneration~(ha^-5)",
    "growth~('%'~yr^-1)",
    "mortality~('%'~yr^-1)"
  )
  p_dat2[,variable2 := factor(
    variable, 
    levels = vars,
    labels = var_labels2),]
  
  
  p_dat2[, pred_obs2 := factor(pred_obs, levels = c("obs","pred"), labels = c("observation","prediction")) ,]
  
  if(grepl("period35",chosen_scale)){
    p_dat2[, year2 := 1984+year,]
  }else{
    p_dat2[, year2 := 1985+year*5,]
  }
  
  p_dat2cast <- dcast(p_dat2, type + year2 + variable2 + species ~ pred_obs2, value.var = "value")
  p_dat2cast[,species2 := factor(species, levels = 1:5, labels = names(pft_cols)),]
  p_dat2[,type2 := type,]
  p_dat2[pred_obs == "obs",type2 := "observation",]
  p_dat2[,type2 := factor(type2, levels = c("observation", "process", "hybrid"), 
                          labels = c("Observed", "Predicted (Process Model)", "Predicted (Hybrid Model)")),]
  p_dat2[is.na(species2), species2 := "all",]
  
  p_list <- list()
  for(j_type in c("process","hybrid")){
    i_pdat = p_dat2[type == j_type,]
    j_linetype = ifelse(j_type == "process", "dashed", "dotted")
    j_pointtype = ifelse(j_type == "process", 1, 2)
   }
  p_dat2cast[is.na(species2),species2 := "all",]
  all_cors[,type2 := factor(type, levels = c("process", "hybrid"), labels = c("Process", "Hybrid")),]
  all_cors[,variable:= factor(variable, levels = c("ba","trees","dbh","reg", "growth", "mort"),ordered = T)]
  all_cors[is.na(species2),species2 := "all",]

}




cors_naive = fread("results/cors_naive.csv")

all_cors
names(cors_naive)[names(cors_naive) == "r"] <- "spearmans_r"
names(cors_naive)[names(cors_naive) == "var"] <- "variable"
cors_naive$type2 = "Naive DNN"
var_labels3 = c("Basal Area [m²/ha]",
                "Trees [N/ha]",
                "DBH [cm]",
                "Regeneration [N/ha]",
                "Growth [%/100]",
                "Mortality [%/100]")
cors_naive[,variable2 := factor(
    variable, 
    levels = vars,
    labels = var_labels3),]
cors_naive[,species2 := factor(species, levels = 1:5, labels = names(pft_cols)),]
cors_naive[is.na(species2), species2 := "all",]

all_cors2 <- rbind(all_cors, cors_naive, fill = T)

var_labels3 = c("Basal Area [m²/ha]",
                "Trees [N/ha]",
                "DBH [cm]",
                "Regeneration [N/ha]",
                "Growth [%/100]",
                "Mortality [%/100]")
all_cors2[,variable2 := factor(
    variable, 
    levels = vars,
    labels = var_labels3),]

p2_full = 
  ggplot(all_cors2, aes(x = type2, y = spearmans_r, color = species2))+
    geom_hline(yintercept = c(-1,0,1), linetype = "solid", color = "grey50", linewidth = 1)+
    geom_boxplot(position = position_dodge())+
    labs(x = "", y = "Spearmans Correlation")+
    theme_classic()+
    facet_wrap(~variable2, ncol = 3)+
    coord_cartesian(ylim = c(-1,1))+
    # fully remove stip
    theme(legend.position = "bottom")+
    #move y axis from left side to right side
    scale_y_continuous(position = "left", breaks = c(-1,0,1))+
    scale_color_manual(name = "PFT", values = c(pft_cols, all = "black")) + guides(colour = guide_legend(nrow = 2))
p2_full

```

#### Table 1 - Simulation vs. Observation

```{r}
# all_cors2[is.na(cv),cv := "S0T0",]
# all_cors2[is.na(test_train),test_train := "test",]
# all_cors2[type2 == "Naive DNN"]
# all_cors2
# dcast(all_cors2[species2 != "all" & test_train == "test" & cv != "S0T0", rmse ,by = .(variable2, type2)], variable2~type2, value.var = "rmse", fun.aggregate = mean)
# dcast(all_cors2[(is.na(k)|k == 3) & species2 != "all" & test_train == "train" & cv != "S0T0", rmse ,by = .(variable2, type2)], variable2~type2, value.var = "rmse", fun.aggregate = mean)
rmse_pdat <- all_cors2[cv != "S0T0", rmse ,by = .(variable2, test_train,type2, species2, cv)]
dcast(rmse_pdat[species2 != "all" & test_train == "test"], variable2~type2+test_train, value.var = "rmse", fun.aggregate = mean)
```


### Figure 4 - Succession trajectories

```{r}
#| label: fig-figure_4
#| fig-format: pdf
#| fig-width: 10
#| fig-height: 6
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Simulated successional trajectories of stand characteristics and demographic rates of five PFTs for Process-FINN (a) and Hybrid-FINN (b), where the growth process is replaced by a DNN. Straight horizontal lines represent the average observed values from the original BCI forest inventory data. Note that rates are not annual and represent changes of five-year intervals. Both models were trained with the BCI forest data with 7 censuses. Forest dynamics were simulated for 1200 years. Disturbance regime of Rüger et al. (2020) was used."

obs_dt <- fread("data/BCI/noSplits/pft-period7-25patches/obs_dt.csv")

obs_dt[,species2 := factor(species,levels = 1:5, labels = names(pft_cols)),]

var_levels <- c("ba", "trees", "dbh", "AL", "growth", "mort", "r_mean_ha")
var_labels <- c(
  "Basal Area [m²/ha]",
  "Trees [N/ha]",
  "avg. DBH [cm]",
  "Available Light [%]",
  "Growth [%/100]",
  "Mortality [%/100]",
  "Regeneration [N/ha]"
)

obs_dt_year <- obs_dt[, .(
  ba        = mean(ba, na.rm = TRUE) / 0.1,
  trees     = mean(trees, na.rm = TRUE) / 0.1,
  dbh       = mean(dbh, na.rm = TRUE),
  growth    = mean(growth, na.rm = TRUE),
  mort      = mean(mort, na.rm = TRUE),
  r_mean_ha = mean(reg, na.rm = TRUE)
), by = .(year, species2)]

avg_obs_dt <- obs_dt_year[, .(
  ba        = mean(ba, na.rm = TRUE),
  trees     = mean(trees, na.rm = TRUE),
  dbh       = mean(dbh, na.rm = TRUE),
  growth    = mean(growth, na.rm = TRUE),
  mort      = mean(mort, na.rm = TRUE),
  r_mean_ha = mean(r_mean_ha, na.rm = TRUE)
), by = .(species2)]

avg_obs_dt_max <- obs_dt_year[, .(
  ba        = max(ba, na.rm = TRUE),
  trees     = max(trees, na.rm = TRUE),
  dbh       = max(dbh, na.rm = TRUE),
  growth    = max(growth, na.rm = TRUE),
  mort      = max(mort, na.rm = TRUE),
  r_mean_ha = max(r_mean_ha, na.rm = TRUE)
), by = species2]

avg_obs_dt_min <- obs_dt_year[, .(
  ba        = min(ba, na.rm = TRUE),
  trees     = min(trees, na.rm = TRUE),
  dbh       = min(dbh, na.rm = TRUE),
  growth    = min(growth, na.rm = TRUE),
  mort      = min(mort, na.rm = TRUE),
  r_mean_ha = min(r_mean_ha, na.rm = TRUE)
), by = species2]

obs_dt_year2 <- melt(obs_dt_year,     id.vars = c("year","species2"), variable.name = "variable", value.name = "value")
avg_obs_dt2 <- melt(avg_obs_dt,     id.vars = "species2", variable.name = "variable", value.name = "value")
avg_obs_dt2_max <- melt(avg_obs_dt_max, id.vars = "species2", variable.name = "variable", value.name = "value")
avg_obs_dt2_min <- melt(avg_obs_dt_min, id.vars = "species2", variable.name = "variable", value.name = "value")

for (dt in list(avg_obs_dt2, avg_obs_dt2_max, avg_obs_dt2_min,obs_dt_year2)) {
  dt[, variable2 := factor(variable, levels = var_levels, labels = var_labels)]
}


## Simulate equilibrium from fitted models
# load fitted BCI models
hybrid_m = torch::torch_load(fitted_models[grepl("hybrid",fitted_models)])
process_m = torch::torch_load(fitted_models[!grepl("hybrid",fitted_models)])

i_name = "pft-period7-25patches"
# # read full raw data
# obs_dt2 = fread(paste0("data/BCI/noSplits/", i_name,"/obs_dt.csv"))
# obs_dt = fread(paste0("data/BCIsecf/stand_dt_secf.csv"))
# cohorts_dt2 = fread(paste0("data/BCI/noSplits/", i_name,"/initial_cohorts1985.csv"))
# cohorts_dt = fread(paste0("data/BCIsecf/initial_cohorts_10patches_0.1ha_age40.csv"))
# cohorts_dt$dbh = cohorts_dt$dbh_cm
env_dt = fread(paste0("data/BCI/noSplits/", i_name,"/env_dt.csv"))
# obs_dt = fread(paste0("data/BCI/noSplits/", i_name,"/obs_dt.csv"))

# predict for model
# cohort1 <- FINN::CohortMat(obs_df = cohorts_dt, sp = uniqueN(obs_dt$species))

# create artificial time series
env_dt_long <- data.table()
start_t = 0
for(k in 1:40){
  env_dt_temp = copy(env_dt)
  # randomly shuffle years in env_dt
  env_dt_temp$year = env_dt_temp$year+start_t
  env_dt_temp[,year := year[sample(length(year))], by = siteID]
  # env_dt_temp$year = env_dt_temp$year[sample(length(env_dt_temp$year), replace = F)]
  env_dt_long = rbind(env_dt_long, env_dt_temp)
  start_t = max(env_dt_long$year)
  # if(start_t > ceiling(80/uniqueN(env_dt$year))) break
}
# env_dt_long$swp = 0
# env_dt_long[,":="(
#   Prec = mean(Prec),
#   SR_kW_m2 = mean(SR_kW_m2),
#   RH_prc = mean(RH_prc),
#   T_max = mean(T_max),
#   T_min = mean(T_min)
# ), by = siteID]
# here we specify a disturbance frequency of 1%, which means that there is a 1% chance each year that a disturbance occurs
disturbance_frequency = 1

Ntimesteps = uniqueN(env_dt_long$year)
Ntimesteps = max(env_dt_long$year)
Nsites = uniqueN(env_dt_long$siteID)
# the disturbance intensity at each timestep is the fraction of patches that is disturbed at that time step
# here we specify a disturbance frequency of 1%, which means that there is a 1% chance each year that a disturbance occurs
disturbance_intensity = runif(Ntimesteps*Nsites, 0.0043, 0.016)
#length(disturbance_intensity)
dist_dt <- env_dt_long[,.(siteID, year)]
# this will result in 0 to 20 % of the patches being disturbed at each timestep with a change of 1% that a disturbance occurs at the timestep
dist_dt$intensity = rbinom(Ntimesteps*Nsites, 1, disturbance_frequency)*disturbance_intensity

pred_hybrid = hybrid_m$simulate(env = env_dt_long, disturbance = dist_dt, init_cohort = NULL, patches = 20, patch_size = 0.1)

pred_process = process_m$simulate(env = env_dt_long, disturbance = dist_dt, init_cohort = NULL, patches = 20, patch_size = 0.1)

naive_preds <- fread("results/naive_succession.csv")

patch_size = 0.1
predictions = list()
predictions[["pred_hybrid"]] = pred_hybrid
predictions[["pred_process"]] = pred_process
rmse_dt_all <- data.table()
p_list = list()

# First pass: collect all prediction data to compute shared axis limits
all_p_dat <- data.table()
for(i in c("pred_hybrid","pred_process")){
  p_dat <- predictions[[i]]$long$site[, .(value = mean(value)), by = .(year = 1+(year-1)*5, species, variable)]
  p_dat$model = i
  p_dat[,species2 := factor(species, levels = 1:5, labels = names(pft_cols)),]
  p_dat[, variable2 := factor(
    variable,
    levels = c("ba", "trees", "dbh", "AL", "growth", "mort", "r_mean_ha"),
    labels = c("Basal Area [m²/ha]", "Trees [N/ha]", "avg. DBH [cm]",
               "Available Light [%]", "Growth [%/100]", "Mortality [%/100]", 
               "Regeneration [N/ha]")
  ),]
  p_dat <- p_dat[!is.na(variable2)]
  p_dat[variable %in% c("ba", "trees"), value := value/patch_size,]
  all_p_dat <- rbind(all_p_dat, p_dat)
}

# Compute global max per variable across both models
global_max <- all_p_dat[, .(ymax = max(value, na.rm = TRUE)), by = variable]

# Second pass: build plots using shared limits
for(i in c("pred_hybrid","pred_process")){
  p_dat <- all_p_dat[model == i]
  
  rmse_dt <- merge(p_dat[year > 750], avg_obs_dt2, by = c("species2", "variable", "variable2"), suffixes = c(".sim",".obs"))
  rmse_dt_all <- rbind(rmse_dt_all, data.table(rmse_dt, type = i))
  
  p_dat_dummy = data.table::copy(p_dat)
  for(v in global_max$variable) {
    p_dat_dummy[variable == v][1:2]$value = c(0, global_max[variable == v, ymax])
  }
  
  p <- ggplot(p_dat, aes(x = year, y = value, color = species2)) +
    geom_line(linewidth = 1, alpha = 0.5) +
    labs(x = "Year", y = "Value") +
    coord_cartesian(ylim = c(0, NA)) +
    facet_wrap(~variable2, scales = "free_y", ncol = 2, strip.position = "left") +
    geom_blank(data=p_dat_dummy) + 
    theme_classic() +
    scale_color_manual(
      name = "PFT", values = pft_cols)+
    theme(
      axis.title.y = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 90),
      strip.background = element_blank(),
      legend.position = "bottom",
      legend.key.size = unit(0.001, "mm"),
      legend.key.width = unit(2, "line")
    )+
    guides(color = guide_legend(ncol = 1,byrow = TRUE))
  legend_p <- get_legend(p)
  
  p_list[[i]] <- p
}

grid.arrange(
  grobs = c(
    list(label_grob_f("a"), label_grob_f("b")),
    list(p_list[[2]]
      +geom_hline(data = avg_obs_dt2,aes(yintercept = value, color = species2), linetype = "solid", linewidth = 0.5)
      # +geom_hline(data = avg_obs_dt2_max,aes(yintercept = value, color = species2), linetype = "dashed", linewidth = 0.5)
      # +geom_hline(data = avg_obs_dt2_min,aes(yintercept = value, color = species2), linetype = "dashed", linewidth = 0.5)
      +theme(legend.position = "none")
      ,
         p_list[[1]]
      +geom_hline(data = avg_obs_dt2,aes(yintercept = value, color = species2), linetype = "solid", linewidth = 0.5)
      # +geom_hline(data = avg_obs_dt2_max,aes(yintercept = value, color = species2), linetype = "dashed", linewidth = 0.5)
      # +geom_hline(data = avg_obs_dt2_min,aes(yintercept = value, color = species2), linetype = "dashed", linewidth = 0.5)
      +theme(legend.position = "none")), list(legend_p))
      ,
  ncol = 2,
  layout_matrix = rbind(c(1, 2), c(3, 4), c(5,5)),
  heights = c(0.05, 0.90,0.2)
)
```

#### Table 1 Succession trajectories

```{r}

results_naive = data.table(read.csv("results/naive_succession.csv")) 
results_naive[, variable2 := factor(
  variable,
  levels = c("ba.x", "trees.x", "dbh.x", "growth.x", "mort.x", "reg.x"),
  labels =  c("Basal Area [m²/ha]", "Trees [N/ha]", "avg. DBH [cm]",
               "Growth [%/100]", "Mortality [%/100]", 
              "Regeneration [N/ha]")
  ),]
results_naive[variable %in% c("ba.x", "trees.x"), value := value/0.1,]

results_naive[,value.sim := value,]

results_naive <- merge(results_naive[,-"value"], unique(rmse_dt_all[,.(variable2,species2, value.obs)]), by = c("species2","variable2"))

rmse_dt_all2 <- rbind(data.table(results_naive[cv != "S0T0"], type = "naive"), rmse_dt_all, use.names = T, fill = T)

# ggplot(rmse_dt_all2[variable == "mort"], aes(x = year, color = species2))+
#   geom_line(aes(y = value.sim, linetype = "sim"))+
#   geom_line(aes(y = value.obs, linetype = "obs"))+
#   facet_wrap(~type)

rmse_dt_all3 <- rmse_dt_all2[(test_train == "test" | is.na(test_train)),.(
  rmse = round(mean(sqrt((value.sim-value.obs)^2), na.rm=T),3)
), by = .(species2, variable2, type)]

dcast(rmse_dt_all3, variable2~type, value.var = "rmse", fun.aggregate = mean)

```


### Figure 5 - Explainable AI

```{r}
#| label: fig-figure_5
#| fig-format: pdf
#| fig-width: 8
#| fig-height: 5
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Simulated relative growth response to available light and tree size (dbh) for 5 PFTs for the hybrid model (a) and the process model (b). Visualizations are based on accumulated local effect plots. Uncertainties for Hybrid-FINN (A) are based on Monte-Carlo Dropout. Swp is the observed soil water potential."

load("results/03_xAI/results_xAI.RData")

pft_cols <- c(
  Slow                         = "#BF7DA5",  # pink-mauve
  Fast                         = "#DBA242",  # ochre-orange
  `Long-lived pioneer (LLP)`   = "#479C77",  # medium green
  `Short-lived breeder (SLB)`  = "#3471AE",  # mid blue
  Intermediate                 = "#C36837"   # terracotta
)

addA = function(col, alpha = 0.25) apply(sapply(col, grDevices::col2rgb)/255, 2, function(x) grDevices::rgb(x[1], x[2], x[3], alpha=alpha))


tmp = data.frame(df_results_ale) |> group_by(features, species, x) |> summarize(features = unique(features), x = mean(x), y = mean(y), species = unique(species)) |> as.data.table()
tmp_sd = data.frame(df_results_ale) |> group_by(features, species, x) |> summarize(features = unique(features), x = mean(x), y = sd(y), species = unique(species)) |> as.data.table()


par(mfrow = c(2, 3), mar = c(1.5, 2.0, 0, 1.0), oma = c(5, 5, 2, 2))
plot(NULL, NULL, ylim = c(0, 0.82), xlim = c(0, 130), las = 1, xlab = "", ylab="growth", xaxt = "n", axes=F)
text(-30, y = 0.8, pos = 3, xpd = NA, cex = 1.5, font = 2, label = "a")
text(srt = 90, label = "growth", x = -30, y = 0.36, pos = 3, xpd = NA)

for(i in 1:5) polygon(x = c(tmp[features=="dbh"][species==i]$x, rev(tmp[features=="dbh"][species==i]$x)),
                      y = c(tmp[features=="dbh"][species==i]$y - 1.96*tmp_sd[features=="dbh"][species==i]$y, rev(tmp[features=="dbh"][species==i]$y + 1.96*tmp_sd[features=="dbh"][species==i]$y)), 
                      lwd = 1.6, 
                      col =  addA(pft_cols[i], alpha = 0.4), border = NA)

for(i in 1:5) points(tmp[features=="dbh"][species==i][,3:4], type = "l", lwd = 1.6, 
                     col =  pft_cols[i])
axis(2, las = 1)

plot(NULL, NULL, ylim = c(0, 0.82), xlim = c(0, 1), las = 1, xlab = "", ylab="", yaxt = "n", xaxt = "n", axes = F)
for(i in 1:5) polygon(x = c(tmp[features=="light"][species==i]$x, rev(tmp[features=="light"][species==i]$x)),
                      y = c(tmp[features=="light"][species==i]$y - 1.96*tmp_sd[features=="light"][species==i]$y, 
                            rev(tmp[features=="light"][species==i]$y + 1.96*tmp_sd[features=="light"][species==i]$y)), 
                      lwd = 1.6, 
                      col =  addA(pft_cols[i], alpha = 0.4), border = NA)
for(i in 1:5) points(as.data.table(tmp)[features=="light"][species==i][,3:4], type = "l", lwd = 1.6, 
                     col =  pft_cols[i])
axis(2, las = 1)

plot(NULL, NULL, ylim = c(0, 0.82), xlim = c(-1.8, 1.8), las = 1, xlab = "", ylab="", yaxt = "n", xaxt = "n", axes = F)
for(i in 1:5) polygon(x = c(tmp[features=="swp"][species==i]$x, rev(tmp[features=="swp"][species==i]$x)),
                      y = c(tmp[features=="swp"][species==i]$y - 1.96*tmp_sd[features=="swp"][species==i]$y, 
                            rev(tmp[features=="swp"][species==i]$y + 1.96*tmp_sd[features=="swp"][species==i]$y)), 
                      lwd = 1.6, 
                      col =  addA(pft_cols[i], alpha = 0.4), border = NA)
for(i in 1:5) points(as.data.table(tmp)[features=="swp"][species==i][,3:4], type = "l", lwd = 1.6, 
                     col =  pft_cols[i])
axis(2, las = 1)

plot(NULL, NULL, ylim = c(0, 0.82), xlim = c(0, 130), las = 1, xlab = "dbh", ylab="growth", axes = F)
text(-30, y = 0.8, pos = 3, xpd = NA, cex = 1.3, font = 2, label = "b")
text(srt = 90, label = "growth", x = -30, y = 0.36, pos = 3, xpd = NA)
for(i in 1:5) points(as.data.table(df_results_ale_process)[features=="dbh"][species==i][,3:4], type = "l", lwd = 1.6, 
                     col =  pft_cols[i])
axis(2, las = 1)
axis(1, las = 1)
text(x = 65, y = -0.2, pos = 1, xpd = NA, label = "dbh")

plot(NULL, NULL, ylim = c(0, 0.82), xlim = c(0, 1), las = 1, xlab = "light", ylab="", yaxt = "n", axes =F)
for(i in 1:5) points(as.data.table(df_results_ale_process)[features=="light"][species==i][,3:4], type = "l", lwd = 1.6, 
                     col =  pft_cols[i])
text(x = 0.5, y = -0.2, pos = 1, xpd = NA, label = "light")
axis(2, las = 1)
axis(1, las = 1)

plot(NULL, NULL, ylim = c(0, 0.82), xlim = c(-1.8, 1.8), las = 1, xlab = "swp", ylab="", yaxt = "n", axes =F)
for(i in 1:5) points(as.data.table(df_results_ale_process)[features=="swp"][species==i][,3:4], type = "l", lwd = 1.6, 
                     col =  pft_cols[i])
text(x = 0.0, y = -0.2, pos = 1, xpd = NA, label = "soil water potential")
legend("topleft", bty = "n", col = pft_cols, lty = 1, lwd = 1.3, legend = names(pft_cols))
axis(2, las = 1)
axis(1, las = 1)


```


## Appendix A

### Figure A1 - RMSE boxplots

```{r}
#| label: figure_A1
#| fig-format: pdf
#| fig-width: 7
#| fig-height: 5
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Root mean squared error (RMSE) of observed and simulated variables for five PFTs from the five-fold blocked spatial cross-validation of all models calibrated/trained on the BCI data. 'Process' refers to Process-FINN as a full process model with fixed functional forms, 'Hybrid' refers to Hybrid-FINN in which the growth process is replaced by a DNN, 'Naïve NN' refers to a single DNN that was trained on the response variables for each time step. The naïve NN received the derived response variables of each previous time step as well as the current environment as input. PFTs are according to (Rüger et al., 2020)."

p2_rmse = 
  ggplot(all_cors2[species != "all"], aes(x = type2, y = rmse, color = species2))+
    # geom_hline(yintercept = c(-1,0,1), linetype = "solid", color = "grey50", linewidth = 1)+
    geom_boxplot(position = position_dodge())+
    labs(x = "", y = "RMSE")+
    theme_classic()+
    facet_wrap(~variable2, ncol = 3, scales = "free_y")+
    # coord_cartesian(ylim = c(-1,1))+
    # fully remove stip
    theme(legend.position = "bottom")+
    #move y axis from left side to right side
    # scale_y_continuous(position = "left", breaks = c(-1,0,1))+
    scale_color_manual(name = "PFT", values = c(pft_cols, all = "black")) + guides(colour = guide_legend(nrow = 2))
p2_rmse

```


### Figure A2 - Simulation vs. Observation
```{r}
#| label: figure_A2
#| fig-format: pdf
#| fig-width: 13
#| fig-height: 10
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Observed and predicted forest structure and demographic rates for five PFTs. a) and b) display the observed values (straight line & crosses) vs the predicted values from the process model (dashed line & circles) and the hybrid model (dotted line & triangles), respectively. c) shows the absolute difference between the observed values and the predictions of the process and hybrid model. d) shows the calculated pearson correlation from the five-fold blocked spatial cross-validation. The first two rows show the observed and predicted values for basal area and tree density. The third and fourth row show the observed and predicted values for diameter at breast height (dbh), regeneration, growth, and mortality."

chosen_scale = "pft-period7-25patches"
fitted_models_fig4 <- fitted_models

## Get the single hybrid model
i_hybrid_model <- fitted_models_fig4[grepl("hybrid", fitted_models_fig4)][1]

## Build process and hybrid fit data
fit_process <- build_model_dt(pt_file = fitted_models_fig4[1])[[1]]
fit_hybrid <- build_model_dt(pt_file = i_hybrid_model)[[1]]

all_dt <- rbind(
  data.table(fit_process[test_train == "train"], type = "process"),
  data.table(fit_hybrid[test_train == "train"], type = "hybrid")
)
period_length <- unique(all_dt[!is.na(period_length)]$period_length)
all_dt <- all_dt[, -c("period_length")]

all_dt2 <- melt(all_dt, id.vars = c("type", "siteID", "year", "species", "test_train"))
all_dt2[grepl("pred", variable), pred_obs := "pred"]
all_dt2[grepl("obs", variable), pred_obs := "obs"]
all_dt2[, variable := gsub(".obs|.pred", "", variable)]

p_dat <- all_dt2[test_train == "train", .(value = mean(value, na.rm = T)),
                 by = .(type, year, species, test_train, pred_obs, variable)]
p_dat2 <- p_dat[!is.na(value)]

p_dat2_all <- rbind(
  p_dat2[variable %in% c("ba", "trees"), .(
    value = sum(value, na.rm = T), species = "all"
  ), by = .(type, year, test_train, pred_obs, variable)],
  p_dat2[variable %in% c("dbh", "reg", "growth", "mort"), .(
    value = mean(value, na.rm = T), species = "all"
  ), by = .(type, year, test_train, pred_obs, variable)]
)
p_dat2 <- rbind(p_dat2, p_dat2_all)

p_dat2[variable %in% c("ba", "trees"), value := value / 0.1]
p_dat2[variable %in% c("mortality", "growth"), value := value / 5 * 100]

## --- Read and prepare naive predictions ---
naive_dt <- fread("results/naive_simulated.csv")
naive_dt <- naive_dt[test_train == "train" & cv == "S0T0"]
setnames(naive_dt, "var", "variable")
naive_dt <- naive_dt[pred_obs == "pred"]
naive_dt[, pred_obs := "pred"]
naive_dt[, type := "naive"]

# Drop the pre-computed "all" rows — they have inconsistent units
naive_dt <- naive_dt[species != "all"]

naive_dt <- naive_dt[, .(value = mean(value, na.rm = T)),
                     by = .(type, year, species, test_train, pred_obs, variable)]

# Recompute "all" species the same way as process/hybrid
naive_all <- rbind(
  naive_dt[variable %in% c("ba", "trees"), .(
    value = sum(value, na.rm = T), species = "all"
  ), by = .(type, year, test_train, pred_obs, variable)],
  naive_dt[variable %in% c("dbh", "reg", "growth", "mort"), .(
    value = mean(value, na.rm = T), species = "all"
  ), by = .(type, year, test_train, pred_obs, variable)]
)
naive_dt <- rbind(naive_dt, naive_all)

# Apply same unit conversions as process/hybrid
naive_dt[variable %in% c("ba", "trees"), value := value / 0.1]
naive_dt[variable %in% c("growth"), value := value / 5 * 100]

# Observations are shared with process model
naive_obs <- p_dat2[type == "process" & pred_obs == "obs"]
naive_obs[, type := "naive"]

p_dat2 <- rbind(p_dat2, naive_dt[, .(type, year, species, test_train, pred_obs, variable, value)])
p_dat2 <- rbind(p_dat2, naive_obs)

## --- Shared formatting ---
vars <- c("ba", "trees", "dbh", "reg", "growth", "mort")
var_labels2 <- c(
  "basal~area~(m^2~ha^-1)", "trees~(ha^-1)", "dbh~(cm)",
  "regeneration~(ha^-5)", "growth~('%'~yr^-1)", "mortality~('%'~yr^-1)"
)

p_dat2[, variable2 := factor(variable, levels = vars, labels = var_labels2)]
p_dat2[, species2 := factor(species, levels = 1:5, labels = names(pft_cols))]
p_dat2[is.na(species2), species2 := "all"]
p_dat2[, pred_obs2 := factor(pred_obs, levels = c("obs", "pred"),
                             labels = c("observation", "prediction"))]

if (grepl("period35", chosen_scale)) {
  p_dat2[, year2 := 1984 + year]
} else {
  p_dat2[, year2 := 1985 + year * 5]
}

p_dat2[, type2 := type]
p_dat2[pred_obs == "obs", type2 := "observation"]
p_dat2[, type2 := factor(type2,
  levels = c("observation", "process", "hybrid", "naive"),
  labels = c("Observed", "Predicted (Process Model)",
             "Predicted (Hybrid Model)", "Predicted (Naive Model)")
)]

## --- Legend basis plot ---
cors_fullPlot <- fread("results/cors_fullPlot.csv")
cors_fullPlotSpecies <- fread("results/cors_fullPlotSpecies.csv")
all_cors <- rbind(data.table(cors_fullPlot, species = "all"), cors_fullPlotSpecies)
all_cors <- all_cors[scale == chosen_scale & grepl("T0", cv)]
all_cors[data == strsplit(i_hybrid_model, "/")[[1]][2], type := "hybrid"]
all_cors[data == "02_realdata", type := "process"]
all_cors[, type := factor(type, levels = c("process", "hybrid"), ordered = T)]
all_cors <- all_cors[!is.na(type)]
all_cors[, species := factor(species, levels = sort(unique(all_cors$species)))]
all_cors[, species2 := factor(species, levels = 1:5, labels = names(pft_cols))]

p1_legende_basis <-
  ggplot(p_dat2, aes(x = year2, y = value)) +
  geom_point(aes(shape = type2, color = factor(species2)), size = 3) +
  geom_line(aes(linetype = type2, color = factor(species2))) +
  scale_shape_manual(values = c(4, 1, 2, 0)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash")) +
  scale_color_manual(name = "PFT", values = c(pft_cols, all = "black")) +
  theme_classic() +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    strip.text.y = element_text(angle = 0),
    strip.text.x = element_blank(),
    legend.position = "bottom",
    legend.title.position = "left"
  ) +
  guides(
    shape = guide_legend(position = "bottom", direction = "vertical",
                         title = "Observation / Model Type"),
    linetype = guide_legend(position = "bottom", direction = "vertical",
                            title = "Observation / Model Type")
  )

## --- Helper to build one panel ---
make_panel <- function(j_type, title) {
  i_pdat <- p_dat2[type == j_type]
  ggplot(i_pdat, aes(x = year2, y = value)) +
    geom_point(aes(shape = pred_obs2, color = factor(species2)), size = 3) +
    geom_point(data = p_dat2[type != j_type], aes(x = year2, y = value), color = NA) +
    geom_line(aes(linetype = pred_obs2, color = factor(species2))) +
    scale_shape_manual(values = c(4, 1)) +
    scale_linetype_manual(values = c("solid", "dashed")) +
    facet_wrap(~ variable2, scales = "free_y", ncol = 1,
               strip.position = "left", labeller = label_parsed) +
    scale_color_manual(name = "PFT", values = c(pft_cols, all = "black")) +
    theme_classic() +
    labs(x = "Inventory Period", y = NULL, title = title) +
    theme(
      strip.placement = "outside",
      strip.background = element_blank(),
      strip.text.y = element_text(angle = 0),
      strip.text.x = element_blank(),
      axis.title.x = element_blank(),
      legend.position = "none",
      plot.title = element_text(hjust = 0.5)
    ) +
    guides(
      shape = guide_legend(title = element_blank(), direction = "vertical"),
      linetype = guide_legend(title = element_blank(), direction = "vertical")
    ) +
    scale_x_continuous(breaks = seq(min(i_pdat$year2), max(i_pdat$year2), by = 5))
}

## --- Build the three panels ---
p_process <- make_panel("process", "Process Model")
p_hybrid  <- make_panel("hybrid", "Hybrid Model")
p_naive   <- make_panel("naive", "Naive DNN")

## --- Legend basis (simplified: only obs vs pred) ---
p1_legende_basis <-
  ggplot(p_dat2, aes(x = year2, y = value)) +
  geom_point(aes(shape = pred_obs2, color = factor(species2)), size = 3) +
  geom_line(aes(linetype = pred_obs2, color = factor(species2))) +
  scale_shape_manual(values = c(4, 1)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_color_manual(name = "PFT", values = c(pft_cols, all = "black")) +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.title.position = "left"
  ) +
  guides(
    shape = guide_legend(position = "bottom", direction = "vertical", title = ""),
    linetype = guide_legend(position = "bottom", direction = "vertical", title = "")
  )

p1to3_legend <- get_legend(p1_legende_basis)

gridExtra::grid.arrange(
  grobs = list(
    p_process, p_hybrid, p_naive,
    p1to3_legend
  ),
  layout_matrix = rbind(
    c(1, 2, 3),
    c(4, 4, 4)
  ),
  widths = c(0.6, 0.6, 0.6),
  heights = c(0.95, 0.1)
)

```

### Figure A3 - Emergent patterns
```{r}
#| label: figure_A3
#| fig-format: png
#| fig-width: 9
#| fig-height: 8
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Annual growth (b) and mortality (a) rates of trees as a function of stem diameter (dbh), shown separately for the hybrid and process models (rows) and the five plant functional types (columns). Grey hexagonal bins show the density of simulated cohort-level values (shading scaled to the number of simulated cohorts per bin); lines are generalized additive models (GAMs) fitted to the observed (red) and simulated (cyan) data, summarising how well the simulated size-dependent patterns match the observed ones. Rates are expressed as annual proportions (yr-1). Note that the observed mortality data consist of binary outcomes (0 = survived, 1 = died) aggregated per PFT, whereas the simulated values are estimated annual mortality probabilities; the observed data are therefore represented only by the fitted GAM and not as hexbins."
# load models
process_m <- torch::torch_load(fitted_models[1])
hybrid_m <- torch::torch_load(fitted_models[2])
if(file.exists("figures/emergent-patterns.csv")){
  p_dat <- fread("figures/emergent-patterns.csv")
}else{
  cohort_dt <- fread("data/BCI/noSplits/pft-period7-25patches/initial_cohorts1985.csv")
  env_dt <- fread("data/BCI/noSplits/pft-period7-25patches/env_dt.csv")
  cohort1 <- CohortMat$new(obs_df = cohort_dt)
  plot(process_m$parameters_r$par_regeneration_unconstrained)
  obs_trees <- fread("data/BCI/data-cleaning/pft/all_trees.csv")

  proc_sim = process_m$simulate(env = env_dt, init_cohort = cohort1, patches = 25, patch_size = 0.1, debug = T)
  hybr_sim = hybrid_m$simulate(env = env_dt, init_cohort = cohort1, patches = 25, patch_size = 0.1, debug = T)
  proc_dat = data.table(proc_sim$wide$cohort, type = "process")
  hybr_dat = data.table(hybr_sim$wide$cohort, type = "hybrid")
  obs_dat = obs_trees[census > 1985 & status2 != "regeneration",.(
    species,
    g = relative_growth_5yr,
    m = as.integer(status2 == "died"),
    dbh = gdbh_cm_before,
    dbh2 = gdbh_cm
    )]
  # obs_dat <- fread("data/BCI/noSplits/pft-period7-25patches/obs_dt.csv")

  p_dat <- rbindlist(list(
      data.table(proc_dat, dbh2 = proc_dat$dbh, obs = "simulated"),
      data.table(hybr_dat, dbh2 = hybr_dat$dbh, obs = "simulated"),
      data.table(obs_dat, obs = "observed", type = "hybrid"),
      data.table(obs_dat, obs = "observed", type = "process")
      ), fill = T
      )
  fwrite(p_dat, "figures/emergent-patterns.csv")
}
p_dat[,species2 := factor(species, levels = 1:5, labels = names(pft_cols)),]
p1 = ggplot()+
  geom_hex(
    data = p_dat[dbh >= 1 & obs != "observed"], aes(dbh,(1+m)^(1/5)-1), bins = 30)+
  scale_fill_gradient(
    name = "No. of simulated cohorts", low = "grey85", high = "grey15", trans = "log10")+
  geom_smooth(
    data = p_dat[dbh >= 1], aes(dbh,(1+m)^(1/5)-1, color = obs),
    method = "gam", method.args = list(family = "binomial")
    )+
  facet_grid(factor(type, levels = c("observed", "hybrid", "process"), ordered = T)~species2)+
  scale_x_log10()+
  theme_classic()+
  labs(x = "dbh [cm]", y = "Annual mortality rate")
dbh_cmTOba_m2 <- function(dbh) {
  dbh = dbh/100
  return(pi*dbh^2/4)
  }
p2 = ggplot()+
  geom_hex(
    data = p_dat[dbh2 >= 1 & g > 0 & (m != 1|is.na(m)) & obs != "observed"],
    aes(dbh,(1+g)^(1/5)-1), bins = 30)+
  scale_fill_gradient(
    name = "No. of simulated cohorts", low = "grey85", high = "grey15", trans = "log10")+
  geom_smooth(
    data = p_dat[dbh2 >= 1 & g > 0 & (m != 1|is.na(m))], aes(dbh,(1+g)^(1/5)-1, color = obs),
    method = "gam", method.args = list(family = "gaussian")
    )+
  coord_cartesian(ylim = c(0,NA))+
  facet_grid(factor(type, levels = c("observed", "hybrid", "process"), ordered = T)~species2)+
  scale_x_log10()+
  theme_classic()+
  labs(x = "dbh [cm]", y = "Annual growth rate")
leg_fill  <- get_legend(p1 + theme(legend.position = "bottom"))
leg_color <- get_legend(p2 + theme(legend.position = "bottom", legend.title = element_blank()))
legends   <- arrangeGrob(grobs = list(leg_fill, leg_color), ncol = 2)
grid.arrange(grobs = list(
  p1+ggtitle("a")+theme(legend.position = "none"),
  p2+ggtitle("b")+theme(legend.position = "none"),
  legends
  ),
  layout_matrix = rbind(
    c(1),
    c(2),
    c(3)
  ), widths = c(1), heights = c(0.5,0.5,0.12))

```
### Figure A4 - Fitted process parameters

```{r}
#| label: figure_A4
#| fig-format: pdf
#| fig-width: 10
#| fig-height: 8
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Estimated process-level parameters of Process-FINN (red) and Hybrid-FINN (blue) calibrated on the BCI forest inventory data, grouped by demographic process (competition, growth, mortality, regeneration) and shown for each of the five plant functional types. These are the mechanistic parameters of the model's process equations: competition (compHeight, compStr), growth (growthLight, growthSize), mortality (mortLight, mortSize, mortGrowth), and regeneration (regLight). Because Hybrid-FINN replaces the growth process with a DNN, it has no mechanistic growth parameters; the growth panels therefore show estimates for Process-FINN only. The parameters of the environmental-response functions are not shown here; they are reported separately in Figure A5."

models_process = list.files("results/02_realdata/", full.names = T, recursive = T, pattern = "pft-period7-25patches_S0_T0")
models_hybrid = list.files("results/02_realdata_hybridTF0dropout0.1/", full.names = T, recursive = T, pattern = "pft-period7-25patches_S0_T0")
m_true_process = torch::torch_load(models_process)
m_true_hybrid = torch::torch_load(models_hybrid)

par_df = data.table()
par_names = c("par_competition_r", "par_growth_r", "par_mortality_r", "par_regeneration_r")
for(i in par_names){
  for(c in 1:ncol(m_true_process[[i]])){
    par_df = rbind(
      par_df,
      data.table(
        true_process = m_true_process[[i]][,c],
        true_hybrid = m_true_hybrid[[i]][,c],
        par = paste0(i,c), 
        species = 1:length(m_true_process[[i]][,c])
        )
      )
  }
}

par_names_nn = c("nn_mortality.0.weight", "nn_growth.0.weight", "nn_regeneration.0.weight")
for(i in par_names_nn){
  for(c in 1:ncol(m_true_process$parameters_r[[i]])){
    par_df = rbind(
      par_df,
      data.table(
        true_process = m_true_process$parameters_r[[i]][,c], 
        true_hybrid = m_true_hybrid$parameters_r[[i]][,c], 
        par = paste0(i,c), 
        species = 1:length(m_true_process$parameters_r[[i]][,c])
        ), fill = TRUE
    )
  }
}

par_df[grepl("competition",par),process := "competition",]
par_df[grepl("competition_r1",par),name := "compHeight",]
par_df[grepl("competition_r2",par),name := "compStr",]
par_df[grepl("growth",par),process := "growth",]
par_df[grepl("growth_r1",par),name := "growthLight",]
par_df[grepl("growth_r2",par),name := "growthSize",]
par_df[grepl("mortality",par),process := "mortality",]
par_df[grepl("mortality_r1",par),name := "mortLight",]
par_df[grepl("mortality_r2",par),name := "mortSize",]
par_df[grepl("mortality_r3",par),name := "mortGrowth",]
par_df[grepl("regeneration",par),process := "regeneration",]
par_df[grepl("regeneration_r1",par),name := "regLight",]
par_df[grepl("par",par),type := "process parameter",]
par_df[grepl("nn_",par),type := "environment parameter",]

par_df[type == "environment parameter",name := gsub("weight","",tstrsplit(par, ".", fixed = T)[[3]]),]

# make "intercept" allways the first
par_df[,name := factor(name, levels = c("intercept", unique(par_df[name != "intercept"]$name))),]

p_dat <- melt(par_df, id.vars = c("name", "process", "species", "type"), 
                 measure.vars = c("true_process", "true_hybrid"), 
                 variable.name = "true", value.name = "value")

p_dat[,model := gsub("true_","",true),]

p_dat[,species2 := factor(species, levels = 1:5, labels = names(pft_cols)),]

p_proc_comp <- ggplot(
  p_dat[type == "process parameter" & process == "competition"], 
  aes(x = name, y = value, color = factor(model), fill = factor(model))
  ) +
  geom_hline(yintercept = 0)+
  facet_wrap(process~paste0(species2), ncol = 5, scales = "fixed") +
  theme_classic()+
  geom_bar(width = 0.2, outlier.shape = NA, stat = "identity", position = position_dodge())+
  theme(axis.title.x = element_blank())
p_proc_growth <- ggplot(
  p_dat[type == "process parameter" & process == "growth" & model == "process"], 
  aes(x = name, y = value, 
      color = factor(model, levels = c("hybrid", "process")), 
      fill = factor(model, levels = c("hybrid", "process")))
  ) +
    geom_hline(yintercept = 0)+
  scale_color_manual(values = c("process" = "#00BFC4"))+
  scale_fill_manual(values = c("process" = "#00BFC4"))+
  facet_wrap(process~paste0(species2), ncol = 5, scales = "fixed") +
  theme_classic()+
  geom_bar(width = 0.2, outlier.shape = NA, stat = "identity", position = position_dodge())+
  theme(axis.title.x = element_blank())
p_proc_mort <- ggplot(
  p_dat[type == "process parameter" & process == "mortality"], 
  aes(x = name, y = value, color = factor(model), fill = factor(model))
  ) +
    geom_hline(yintercept = 0)+
  facet_wrap(process~paste0(species2), ncol = 5, scales = "fixed") +
  theme_classic()+
  geom_bar(width = 0.2, outlier.shape = NA, stat = "identity", position = position_dodge())+
  theme(axis.title.x = element_blank())
p_proc_reg <- ggplot(
  p_dat[type == "process parameter" & process == "regeneration"], 
  aes(x = name, y = value, color = factor(model), fill = factor(model))
  ) +
    geom_hline(yintercept = 0)+
  facet_wrap(process~paste0(species2), ncol = 5, scales = "fixed") +
  theme_classic()+
  geom_bar(width = 0.2, outlier.shape = NA, stat = "identity", position = position_dodge())+
  theme(axis.title.x = element_blank())

p_proc_legend <- get_legend(p_proc_comp+theme(legend.position = "bottom")+guides(color = guide_legend(title = "Model"), fill = guide_legend(title = "Model"))+guides(color = "none"))
# arrange in three columns

fac_levels = c("intercept","Prec","SR_kW_m2","RH_prc","T_max","T_min","swp")

p_env <- ggplot(
  p_dat[type == "environment parameter"],
       aes(x = factor(name, levels = 1:7, labels = fac_levels), y = value, color = factor(model), fill = factor(model))
  ) +
  geom_hline(yintercept = 0)+
  facet_wrap(process~paste0(species2), ncol = 5, scales = "free_x") +
  labs(x = "Parameter", y = "Difference")+
  theme_classic()+
  geom_bar(width = 0.7, outlier.shape = NA, stat = "identity", position = position_dodge2(width = 1.5, padding = 0.2, preserve = "single"))+
  theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))


grid.arrange(
  grobs = list(
    p_proc_comp+theme(legend.position = "none", axis.title.y = element_blank()), 
    p_proc_growth+theme(legend.position = "none", axis.title.y = element_blank()), 
    p_proc_mort+theme(legend.position = "none", axis.title.y = element_blank()), 
    p_proc_reg+theme(legend.position = "bottom", legend.title = element_blank(), axis.title.y = element_blank())
    # p_proc_legend
  ), 
  layout_matrix = rbind(
    c(1),
    c(2),
    c(3),
    c(4)
    # c(5)
  ),
  # include letters to each plot
  heights = c(0.6,0.6,0.6,0.7),
)

```

### Figure A5 - Fitted environment parameters

```{r}
#| label: figure_A5
#| fig-format: pdf
#| fig-width: 10
#| fig-height: 8
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Estimated parameters of environmental responses of process-FINN and hybrid-FINN calibrated on the BCI forest inventory data."

p_env+theme(legend.position = "bottom", axis.title.y = element_blank(), legend.title = element_blank())


```

### Figure A6 - Response correlations

```{r}
#| label: figure_A6
#| fig-format: pdf
#| fig-width: 10
#| fig-height: 5
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Spearman correlations from five fold spatial cross validation to asses how well a model can be trained with different combinations of response variables. To ensure a full coverage of each response variable, we simulated new data from the process model that was trained on the BCI forest inventory data. These simulations were used to calibrate models with different combinations of response variables. The correlations were derived as average from a spatially blocked five-fold cross validation."

cors_fullPlot = fread("results/cors_fullPlot.csv")
cors_fullPlotSpecies <- fread("results/cors_fullPlotSpecies.csv")


all_cors <- rbind(
  data.table(cors_fullPlot, species = "all"),
  cors_fullPlotSpecies
)

p_cors <- cors_fullPlot[hybrid == "nohybrid"]
# p_cors <- cors_fullPlot[hybrid == "nohybrid"]

p_cors2 <- p_cors[cv != "T0S0",.(
  r = mean(spearmans_r),
  nresponses = length(tstrsplit(response, ".", fixed = TRUE))
  # nresponses = tstrsplit(response, ".", fixed = TRUE)[[1]]
  ), by = .(variable,response=gsub(".pt","",response),scale, test_train,simreal)]


ggplot(p_cors2[grepl("pft", scale) & simreal == "simulated"],aes(x = variable, y = forcats::fct_reorder(response, -nresponses), fill = r)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limits = c(-1,1)) +
  labs(x = "Predicted Variable", y = "Loss Variables", fill = "Spearmans Correlation") +
  theme_classic() +
  facet_grid(test_train~scale, )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### MISSING Figure A7 - Runtime

### Figure A8 - Naive equilibrium
```{r}
#| label: figure_A8
#| fig-format: pdf
#| fig-width: 7
#| fig-height: 5
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Simulated successional trajectories of stand characteristics and demographic rates of five PFTs for the naive DNN. Note that rates are not annual and represent changes of five year intervals. Model was trained with the BCI forest data with 7 censuses. Forest dynamics were simulated for 600 years."

results_naive = data.table(read.csv("results/naive_succession.csv"))
# results_naive[variable=="ba.x"]$value = results_naive[variable=="ba.x"]$value
# results_naive[variable=="dbh.x"]$value = results_naive[variable=="dbh.x"]$value

# results_naive$year = rep(rep(seq(0, 594, by = 6), each = 5),6)


results_naive[, variable2 := factor(
  variable,
  levels = c("ba.x", "trees.x", "dbh.x", "growth.x", "mort.x", "reg.x"),
  labels =  c("Basal Area [m²/ha]", "Trees [N/ha]", "avg. DBH [cm]",
               "Growth [%/100]", "Mortality [%/100]", 
              "Regeneration [N/ha]")
  ),]

results_naive$variable 

results_naive[variable %in% c("ba.x", "trees.x"), value := value/0.1,]
# results_naive[variable %in% c("ba.x", "trees.x"), value := value/0.1,]
results_naive <- results_naive[cv == "S0T0" & test_train == "test"]

results_naive$year = rep(rep(seq(0, 1194, by = 6), each = 5),6)
results_naive[variable == "growth.x" & value > 2, value := NA,]
  p <- ggplot(results_naive, aes(x = year, y = value, color = species2)) +
    geom_line(linewidth = 1) +
    labs(x = "Year", y = "Value") +
    coord_cartesian(ylim = c(0, NA)) +
    facet_wrap(~variable2, scales = "free_y", ncol = 2, strip.position = "left") +  # Remove label_parsed
     #geom_blank(data=p_dat_dummy) + 
    theme_classic() +
    # ylab("Basal Area [m²/ha]") +
    scale_color_manual(
      name = "PFT", values = pft_cols)+
    theme(
    axis.title.y = element_blank(),
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 90),
    strip.background = element_blank(),
    legend.position = "right",
    legend.key.size = unit(0.001, "mm"),
    legend.key.width = unit(2, "line")
    # legend.position.inside = c(0.9,0.2),
    # legend.background = element_rect(color = "black", fill = NA)
  )+
    guides(color = guide_legend(ncol = 1,byrow = TRUE))
    # ggtitle(paste0("patches = ",gsub("patches_", "", i)))
  legend_p <- get_legend(p)
  

  # pdf("figures/figure_appendix_naive_succession.pdf", width = 7, height = 5 )
  p+geom_hline(data = avg_obs_dt2,aes(yintercept = value, color = species2), linetype = "solid", linewidth = 0.5)
  # dev.off()
```


### Figure A9 - Recovery

```{r}
#| label: figure_A9
#| fig-format: pdf
#| fig-width: 10
#| fig-height: 6
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Parameter recovery from 600 replicate simulations and refits of the full model using five‑fold spatial cross‑validation. Blue half‑violin/half‑box plots show the distribution of estimation errors (posterior mean of the fitted parameter minus its true value). a) Process‑level parameters: errors grouped by ecological process (competition, growth, mortality, regeneration) and shown for the corresponding process parameters (e.g., compHeight, growthLight, mortSize). b) Environmental parameter estimates: errors for the weights associated with the intercept and six climatic covariates (Prec, SR_kW_m2, RH_prc, T_max, T_min, swp) within each process. Because the neural‑network component consists of a single linear layer, these weights are equivalent to classical multivariate linear‑regression coefficients. The red horizontal line indicates perfect recovery (error = 0)."

models = list.files("results/03_full100reps/", full.names = T, recursive = T)
m_true = torch::torch_load("results/01_full/pft-period7-25patches_full.pt")
m_list = sapply(models, torch::torch_load)
m_pars = lapply(m_list, function(x) x$parameters_r)

i = names(m_true$parameters_r)[1]
m_true$nn_growth$parameters

diffs_df = data.frame()
par_names = c("par_competition_r", "par_growth_r", "par_mortality_r", "par_regeneration_r")
for(i in par_names){
  for(c in 1:ncol(m_true[[i]])){
    for(m in m_list[1:length(m_list)]){
      diffs_df = rbind(
        diffs_df,
        data.frame(
          true = m_true[[i]][,c],
          recov = m[[i]][,c],
          diff = m_true[[i]][,c] - m[[i]][,c], 
          par = paste0(i,c)
          )
        )
    }
  }
}

par_names_nn = c("nn_mortality.0.weight", "nn_growth.0.weight", "nn_regeneration.0.weight")
for(i in par_names_nn){
  for(c in 1:ncol(m_true$parameters_r[[i]])){
    for(m in m_list[1:length(m_list)]){
      diffs_df = rbind(
        diffs_df,
        data.frame(
          true = m_true$parameters_r[[i]][,c], 
          recov = m$parameters_r[[i]][,c],
          diff = m_true$parameters_r[[i]][,c] - m$parameters_r[[i]][,c], 
          par = paste0(i,c))
      )
    }
  }
}

log_mod <- function(x, base = exp(1)) {
  # Vorzeichenmatrix (0 bekommt hier schon das richtige Vorzeichen = 0)
  sgn <- sign(x)
  # Logarithmus von (|x| + 1); bei x = 0 ergibt das log(1) = 0
  log_part <- log(abs(x) + 1, base = base)
  # Kombination von Vorzeichen und Betrag
  result <- sgn * log_part
  # Nur zur Sicherheit: stellt sicher, dass exakt x == 0 auch exakt 0 ergibt
  result[x == 0] <- 0
  return(result)
}
log_mod(-10:10)
par(mfrow = c(1,1), mar = c(3,8,1,1))
boxplot(log_mod(diff) ~ par, data = diffs_df, horizontal = T, las = 1, cex.axis = 0.7)
abline(v = 0)

# create transformation for log modulu
library(scales) 
# -----------------------------------------------------------
# 1)  Transformation definieren
# -----------------------------------------------------------
log_mod_trans <- function(base = exp(1)) {
  trans_new(
    name      = paste0("log_mod-", format(base)),
    transform = function(x) sign(x) * log(abs(x) + 1, base = base),
    inverse   = function(y) sign(y) * (base^abs(y) - 1),
    breaks    = pretty_breaks(),      # Tick‑Positionen auf Originalskala
    domain    = c(-Inf, Inf)
  )
}

# # -----------------------------------------------------------
# # 2)  Beispielplot mit log_mod‐Skala, Original‑Labels
# # -----------------------------------------------------------
# ggplot(diffs_df, aes(y = par, x = diff)) +
#   geom_boxplot() +
#   labs(y = "Parameter", x = "Difference") +
#   scale_x_continuous(
#     trans  = log_mod_trans(),           # ← hier anwenden
#     breaks = pretty_breaks(n = 7)       # (optional) schönere Tick‑Abstände
#   ) +
#   theme_minimal()

diffs_dt <- data.table(diffs_df)
diffs_dt[grepl("competition",par),process := "competition",]
diffs_dt[grepl("competition_r1",par),name := "compHeight",]
diffs_dt[grepl("competition_r2",par),name := "compStr",]
diffs_dt[grepl("growth",par),process := "growth",]
# diffs_dt[grepl("nn_growth.0.weight1",par),process := "growth",]
# diffs_dt[grepl("nn_growth.0.weight1",par),name := "intercept",]
diffs_dt[grepl("growth_r1",par),name := "growthLight",]
diffs_dt[grepl("growth_r2",par),name := "growthSize",]
diffs_dt[grepl("mortality",par),process := "mortality",]
# diffs_dt[grepl("nn_mortality.0.weight1",par),process := "mortality",]
# diffs_dt[grepl("nn_mortality.0.weight1",par),name := "intercept",]
diffs_dt[grepl("mortality_r1",par),name := "mortLight",]
diffs_dt[grepl("mortality_r2",par),name := "mortSize",]
diffs_dt[grepl("mortality_r3",par),name := "mortGrowth",]
diffs_dt[grepl("regeneration",par),process := "regeneration",]
diffs_dt[grepl("regeneration_r1",par),name := "regLight",]
# diffs_dt[grepl("nn_regeneration.0.weight1",par),process := "regeneration",]
# diffs_dt[grepl("nn_regeneration.0.weight1",par),name := "intercept",]
# diffs_dt[grepl("par",par) | grepl("weight1", par),type := "process parameter",]
# diffs_dt[grepl("nn_",par) & !grepl("weight1", par),type := "environment parameter",]
diffs_dt[grepl("par",par),type := "process parameter",]
diffs_dt[grepl("nn_",par),type := "environment parameter",]

diffs_dt[type == "environment parameter",name := gsub("weight","",tstrsplit(par, ".", fixed = T)[[3]]),]

# make "intercept" allways the first
diffs_dt[,name := factor(name, levels = c("intercept", unique(diffs_dt[name != "intercept"]$name))),]

p_proc <- ggplot(diffs_dt[type == "process parameter"], aes(x = name, y = diff)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_quasirandom(alpha = 0.1, method = "pseudorandom", color = "#A8C4E0")+
  # geom_violin(fill = NA, color = "black", scale = "width")+
  facet_wrap(~process, ncol = 2, scales = "free") +
  labs(x = "Parameter", y = "Difference")+
  theme_classic()+
  geom_boxplot(width = 0.2, outlier.shape = NA)+
  theme(axis.title.x = element_blank())

fac_levels = c("intercept","Prec","SR_kW_m2","RH_prc","T_max","T_min","swp")

p_env <- ggplot(diffs_dt[type == "environment parameter"], 
       aes(x = factor(name, levels = 1:7, labels = fac_levels), y = diff)
       ) +
  # see::geom_violinhalf(scale = "width", flip = T)+
  geom_hline(yintercept = 0, color = "red") +
  geom_quasirandom(alpha = 0.1, method = "pseudorandom", color = "#A8C4E0")+
  # geom_violin(fill = NA, color = "black", scale = "width")+
  facet_wrap(~process, ncol = 1, scales = "free") +
  labs(x = "Parameter", y = "Difference")+
  theme_classic()+
  geom_boxplot(width = 0.2, outlier.shape = NA)+
  theme(axis.title.x = element_blank())


gridExtra::grid.arrange(
  grobs = list(
    label_grob_f("a"), label_grob_f("b"),
    p_proc, p_env
  ), 
  layout_matrix = rbind(
    c(1,2),
    c(3,4)
  ),
  # include letters to each plot
  widths = c(0.6,0.6), 
  heights = c(0.05,0.95)
)

```


### Figure A10 - RMSE (test vs train)

```{r}
#| label: figure_A10
#| fig-format: pdf
#| fig-width: 12
#| fig-height: 10
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Root mean squared error "
ggplot(all_cors2[cv != "T0S0" & species2 != "all"], aes(x = rmse,y = type2, color = test_train))+
  geom_boxplot(position = position_dodge())+
  facet_wrap(species2~variable2, scales = "free_x")+
  xlab("RMSE")+
  ylab("")+
  theme_classic()+
  theme(legend.title = element_blank())
```

## Appendix C

### Figure C1 - Competition

```{r}
#| label: figure_B1
#| fig-format: pdf
#| fig-width: 10
#| fig-height: 10
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Visualization of the functions for allometries and competition."

ba_dt = 
  data.table(
    expand.grid(
      list(
        trees = 1:1000,
        dbh = seq(0, 100,1)
      )
    )
  )

ba_dt$rowID = 1:nrow(ba_dt)
ba_dt[, ba := BA_stand(dbh, trees,1), by = rowID]

p_ba =
  ggplot(ba_dt, aes(x = dbh, y = trees, fill = ba)) +
  geom_tile() +
  scale_fill_gradientn(
    colours = rev(viridis::viridis(100)),
    limits = c(0, 100),
    na.value = rev(viridis::viridis(100))[100],
    breaks = seq(0, 100, 20),
    labels = c(seq(0, 80, 20),">100"),
    name = "Basal Area\n[m2/ha]"
  )+
  xlab("dbh [cm]")+
  ylab("trees [ha]")+
  theme_classic()

height_dt =
  data.table(
    expand.grid(
      list(
        dbh = seq(0, 300,1),
        parHeight = seq(0,1,0.2)
      )
    )
  )

height_dt$rowID = 1:nrow(height_dt)
height_dt[, height := height(dbh, parHeight), by = rowID]

p_height <-
  ggplot(height_dt, aes(x = dbh, y = height, color = factor(parHeight, ordered = T)))+
  ylab("height(dbh,parHeight)")+
  xlab("dbh [cm]")+
  geom_line()+
  theme_classic()+
  scale_color_discrete(name = "parHeight")

## Competition
parHeight_vec = c(0.7)
parCompStr = c(0.1,0.2,0.3)
parComp = cbind(parHeight_vec, parCompStr)

cohort_layers_dt <- data.table(
  name = c("understory", "midstory", "overstory"),
  trees = c(1000, 300, 50),
  dbh = c(5, 30, 90),
  dbh_shape = c(2, 4, 5)
)

cohort_layers1_dt <- data.table()
for(i_c in 1:nrow(cohort_layers_dt)){
  dbh = cohort_layers_dt$dbh[i_c]
  trees = cohort_layers_dt$trees[i_c]
  dbh_shape = cohort_layers_dt$dbh_shape[i_c]
  temp_cohort_df1 = data.table(FINN::rweibull_cohorts(trees = trees, dbh_shape = dbh_shape, dbh_scale = dbh, dbh_class_range = 0.1, species = 1, siteID = 1))
  temp_cohort_df1$cohortID = paste0(i_c,"_",temp_cohort_df1$cohortID)
  temp_cohort_df1$trees_ha = trees
  temp_cohort_df1$mean_dbh = dbh
  temp_cohort_df1$dbh_shape = dbh_shape
  temp_cohort_df1$name = cohort_layers_dt$name[i_c]
  cohort_layers1_dt <- rbind(cohort_layers1_dt,temp_cohort_df1)
}

cohort_layers2_dt <- data.table()
counter = 0
for(sp in 1:nrow(parComp)){
  counter = counter + 1
  temp_cohort_df2 = copy(cohort_layers1_dt)
  temp_cohort_df2$siteID = counter
  temp_cohort_df2$species = sp
  temp_cohort_df2$parCompStr = parCompStr[sp]
  cohort_layers2_dt <- rbind(cohort_layers2_dt, temp_cohort_df2)
}

cohort_layers2_dt[,cohortID := as.integer(as.factor(cohortID)), by = siteID]
cohort_layers2_dt$cohortID = as.integer(cohort_layers2_dt$cohortID)
cohort_layers2_dt = cohort_layers2_dt[order(cohortID)]

cohort = CohortMat$new(obs_df = cohort_layers2_dt)
basal_area = BA_stand(cohort$dbh, cohort$trees, patch_size_ha = 1)

light = competition(dbh = cohort$dbh, species = cohort$species,
                    trees = cohort$trees, parComp = torch::torch_tensor(parComp),
                    h = NULL, patch_size_ha = 1)

for(i_site in unique(cohort_layers2_dt$siteID)){
  cohort_layers2_dt[siteID == i_site, basal_area2 := torch::as_array(basal_area)[i_site,1,1:nrow(cohort_layers2_dt[siteID == i_site])]]
  cohort_layers2_dt[siteID == i_site, light2 := torch::as_array(light)[i_site,1,1:nrow(cohort_layers2_dt[siteID == i_site])]]
}

cohort_layers2_dt[,":="(ba = round(sum(basal_area2))), by = .(siteID, name)]
cohort_layers2_dt[,":="(dbh_density = dweibull(dbh,shape = dbh_shape, scale = mean_dbh)), by = .(name)]
cohort_layers2_dt[,name := factor(name, levels = c("understory", "midstory", "overstory"), ordered = T),]

label_dt <- cohort_layers2_dt[,.(mean_dbh = mean(dbh), ba = round(sum(basal_area2)), max_density = max(dbh_density)), by = .(siteID, name)]

p_cohorts = ggplot(cohort_layers2_dt)+
  geom_histogram(aes(x = dbh, fill = name, y = ..density..), alpha = 0.3)+
  geom_line(aes(x = dbh, y = dbh_density, color = name), linewidth = 1)+
  # facet_wrap(~siteID)+
  geom_text(data = label_dt, aes(x = mean_dbh, y = max_density+0.03, label = paste0(name,"\n"," ba = ", ba), color = name), size = 5)+
  guides(
    color = guide_legend(title = "layer", override.aes = list(label = "")),
    fill = guide_legend(title = "layer")
    )+
  xlab("dbh [cm]") +
  theme_bw()+
  coord_cartesian(ylim = c(0, max(label_dt$max_density)+0.05))+
  theme(legend.position = "none", axis.text.y = element_blank(), axis.ticks.y = element_blank())

p_light = ggplot(cohort_layers2_dt)+
  geom_boxplot(aes(x = name, y = light2, color = name), linewidth = 1)+
  facet_wrap(~factor(siteID, levels = 1:3, labels = paste("parCompStr =", c(parComp[,2]))))+
  # add axis labels
  xlab("layer") +
  ylab("light [%]") +
  theme_classic()+
  theme(legend.position = "none", axis.title.x = element_blank())

gridExtra::grid.arrange(
  grobs = list(
    label_grob_f("a"), label_grob_f("b"), label_grob_f("c"), label_grob_f("d"),
    p_ba, p_height, p_cohorts, p_light
  ), 
  layout_matrix = rbind(
    c(1,2),
    c(5,6),
    c(3,NA),
    c(7,7),
    c(4,NA),
    c(8,8)
  ),
  # include letters to each plot
  widths = c(0.6,0.6), 
  heights = c(0.05,0.95,0.05,0.95,0.05,0.95)
)


```


### Figure C2 - Scaled sigmoid

```{r}
#| label: figure_S2
#| fig-format: pdf
#| fig-width: 11
#| fig-height: 5
#| echo: false
#| silent: true
#| message: false
#| warning: false
#| fig-cap: "Comparison of light response in Process-FINN for regeneration (a) and growth response (b)."

light = seq(0,1,0.01)
regLight = seq(0.1, 0.9, 0.2)
growthLight = regLight
regLight_f <- function(light,regLight, k = 10) {
  regP = (1 / (1 + exp(-k * (light - regLight))) - 1 / (1 + exp(k * regLight))) / (1 - 1 / (1 + exp(k * (1 - regLight))))
  return(regP)
  }
growthShade_f <- function(light, growthLight, k = 10) {
  shade = ((1 / (1 + exp(-k * (light - growthLight))) - 1 / (1 + exp(k * growthLight))) /
             (1 / (1 + exp(-k * (1 - growthLight))) - 1 / (1 + exp(k * growthLight))))
  return(shade)
  }

reg_dt <- 
  data.table(
      process = "regeneration",
      light = rep(light, length(regLight)),
      regLight = factor(rep(regLight, each = length(light)), ordered = T),
      response = as.vector(sapply(regLight, function(x) regLight_f(light, x), simplify = TRUE))
    )
growth_dt <- data.table(
      process = "growth",
      light = rep(light, length(growthLight)),
      growthLight = factor(rep(growthLight, each = length(light)), ordered = T),
      response = as.vector(sapply(growthLight, function(x) growthShade_f(light, x), simplify = TRUE))
    )

reg_p <- ggplot(reg_dt, aes(light, response, colour = regLight)) +
  geom_line(size = 1) +
  labs(x = "Light",
       y = "Scaled light response of (regeneration)") +
  theme_classic()
# reg_p
growth_p <- ggplot(growth_dt, aes(light, response, colour = growthLight)) +
  geom_line(size = 1) +
  labs(x = "Light",
       y = "Scaled light response of (growth)") +
  theme_classic()
# growth_p

gridExtra::grid.arrange(reg_p+ggtitle("a"), growth_p+ggtitle("b"), ncol = 2)
```
